Tree establishment in models of forest dynamics: a key priority for further research

### 2023-01-27 manuscript/manuscriptRW.Rmd
added changes based on co-authors comments in the M&M section (not finnished yet)

Most of the models overestimate ingrowth rates, but this is not always
compensated by high simulated mortality rates in the early stages of
population dynamics. The effect of light availability on ingrowth rates
is captured better than the effect of temperature and soil moisture, but
patterns are not consistent across models. Simulated species diversity
of ingrowth matches the observed ranges quite well. Models simulating
higher species diversity at the stand level do not feature higher
ingrowth diversity. The ingrowth of *Abies alba* is captured across
temperature and soil moisture gradients, whereas ingrowth of *Fagus
sylvatica* is generally underestimated. *Picea abies* ingrowth is
generally overestimated, particularly at colder sites. Increasing
complexity in the tree establishment modules of the models is not
related to higher accuracy of simulated tree establishment. Furthermore,
individual model design is more important than scale (stand, landscape,
global) and approach (empirical, process-based) to accurately capture
tree establishment.

@@

In the present study, for the first time, a large number of forest
models commonly used to assess forest dynamics under climate change are
evaluated against a continental-scale, multi-species harmonized data set
on tree establishment [@käber2022]. This is a challenging experiment for
these models as they do also have to estimate the forest stand above the
tree establishment. By tree establishment, we refer to the passing of
trees across a specific diameter threshold ("ingrowth"). We included
models that are based on a range of "philosophies", from models
operating at the stand to the global scale as well as the range of
models from empirically derived to "process-based" [e.g., @bugmann1996;
@fabrika2012; @mette2014; @lexer2001]. Because of the large variability
in tree establishment patterns in nature and the large number of factors
driving this process --- including some that are not incorporated in
most models such as deer browsing --- we do not aim at a detailed
statistical evaluation of each model. Instead, we aim at evaluating the
general ingrowth patterns and magnitudes simulated by the models, and
benchmark the simulated regeneration niche of multiple species against
empirical data along a wide environmental gradient of temperature,
moisture, and light availability.

@@

Fifteen models of forest dynamics (including two model featuring
variants) were used to assess tree establishment. They simulate forest
dynamics at the stand, landscape and global scale (Table
\@ref(tab:tableModels)). The approaches used for model construction and
their origin differ strongly, with most of the models featuring a
largely "process-based" approach, whereas two models are based on
formulations derived from the statistical analysis of inventory data
(SYBILA and xComp). Some of the models largely rely on the original
approach underlying forest gap models (e.g., ForClim 1 and LandClim),
whereas some are based on plant ecophysiological processes (e.g.,
FORMIND and iLand). The different approaches underlying the models have
strongly influenced the formulation of tree establishment.

The establishment modules can be categorized as 'regeneration' or
'recruitment' models [@vanclay1994; @konig_tree_2022] (see Table
\@ref(tab:tableModels)). Regeneration models include processes such as
flowering and pollination, seed production, seed dispersal, germination
seedling growth, which ultimately lead to the simulated number of trees.
Recruitment models, in contrast, introduce a certain number of new trees
with certain characteristics such as biomass or diameter, without
explicitly considering earlier development processes. We can further
distinguish models that feature a feedback in the establishment, i.e.,
where the level and species composition of establishment is influenced
by the existing adult tree community via the production of seeds,
seedlings or saplings, from models that do not contain such feedback.

The starting point for in the models differs as well, ranging from
seedbank, seed or seedling to sapling (i.e., trees much larger than 10
cm height, often at least 2-3 m tall). This implies that the larger the
size threshold, the larger the number of ecological processes that must
be aggregated via parameterizations because they cannot be treated
explicitly. This reduces model complexity but comes at the cost of
blurring process representation.

The overall complexity in the establishment modules also varies
considerably. Following @bugmann2022, we can classify seven models
(iLand, PICUS, LandClim, ForCEEPS, LPJ-GUESS, ForClim and TreeMig) as
having rather high complexity in their establishment modules (average
complexity across all processes \>0, Table \@ref(tab:tableModels)). In
most models, the two processes in the establishment modules with the
highest complexity are the statistical approach used to define the
successful number of new trees and how frost occurrence is taken into
account.

Two models, ForClim and ForCEEPs, were used here with two alternative
variants. ForClim variant 1 [@bugmann1996] is based on a recruitment
module that adheres closely to the concept introduced by @botkin1972,
whereas ForClim variant 11 is adopting a rather different approach that
considers space filling along with different 'points of entry' for the
action of specific environmental factors [@huber2020]. These two model
variants allow us to evaluate a more process-based module (variant 11)
against a simple module (variant 1), while the rest of the model
structure is identical. Similarly, the two ForCEEPS variants allow us to
isolate the importance of the canopy feedback via seed trees for the
quantity and quality of simulated ingrowth, as one variant includes this
feedback whereas the other does not.

@@

Ingrowth data covering a wide range of environmental conditions are hard
to obtain, and this is one of the reasons why most models of forest
dynamics have never been confronted with a dataset covering such
gradients over a large number sites, to evaluate how well ingrowth is
captured. The observations used here are from a novel and unprecedented
network of sites in forest reserves that represent the range of
environmental gradients in temperature and precipitation in central
Europe as compiled in the framework of the EuFoRIa network
[@euforia2019] (Figure \@ref(fig:map)). All these forests have not been
managed for at least 10 years prior to their designation as reserves,
and most of them provide time series of natural forest dynamics over
multiple decades and up to ca. 80 years. The census periods range from 3
to 37 years, with an average of 14 years. The data provide information
at the tree level, thus allowing for sequential comparison such as
individual-tree ingrowth and death between the measurements.

We selected 200 sites from this network as the benchmarking data set for
the simulation. The selection of sites is representative for the
environmental variation of the data. This was achieved by applying
kmeans clustering to define 200 clusters of plots from the original set
of 869 plots along the environmental dimensions of temperature, water
balance, soil quality and exposition (i.e., slope and aspect). Each of
these sites featured at least two consecutive measurements. Ingrowth was
measured at threshold of 7 or 10 cm of stem diameter at all sites except
for 35 sites where only ingrowth at the 10 cm diameter threshold was
available. Plot size ranged from 0.02 ha to 5.52 ha, and the data were
further processed and aggregated following @käber2022 to provide
ingrowth rates per ha and per decade. The observations used in this
study featured 30.900 established trees, and ingrowth rates per site,
sample and decade ranged from zero to 1246 trees per ha, with a mean of
56 trees per ha. For more details on this unique dataset and the
detailed establishment information, cf. @käber2022.

@@

The overarching goal of the experiments was to assess the regeneration
niche as it arises from empirical data against its representation in a
wide range of models of forest dynamics. We define the regeneration
niche as the passing of a diameter threshold of 7 or 10 cm, respectively
(synonym: ingrowth). To this end, each modeling group was provided with
a detailed protocol (cf. Supplementary Material 1, henceforth "SM") with
instructions how to perform the simulations, providing input variables
on climate and soil conditions, and the expected output variables.
Neither were further site information (except for the data specified
below) nor any data on tree ingrowth provided prior to the simulation.
That is, the models were run in "blind flight" mode.

The input variables were collected and aggregated to be adapted to the
needs of the different models. Time series of climatic variables from
1981 to 2018 were provided in hourly [Era5-land data, Muñoz Sabater
-@muñozsabater2019], daily and monthly resolution [CHELSA data, Karger
et al. -@karger2021]. Some variables required for some models such as
relative humidity or vapor pressure deficit had to be calculated from
these variables. The final strategy how to use the climate data was
different in each model, based on the approach that best suited the
model (Table \@ref(tab:tableModels)). Soil quality data was provided as
a continuous value between 1 and 5 [Soilgrids dataset, Hengl et al.
-@hengl2017, Poggio et al. -@poggio2021]. The protocol also provided the
elevation, slope and aspect for each of the 200 sites
[@asterscienceteam2019], but no other spatial information such as
coordinates was provided, with the exception of iLand, which used the
blurred coordinates to derive highly detailed soil data.

The simulations were run in the absence of natural disturbances. Only
the model LPJ-GUESS had to include HERE LPJ-GUESS PROVIDE ONE SENTENCE
OF WHY YOU INCLUDED DISTURBANCES AND WHAT TYPE. The simulations were set
up to sample species-specific ingrowth rates per decade and per ha in
the equilibrium state of the model, typically entailing a "spin-up" run
(as we did not provide any forest data). The modeling teams decided on
the simulated area and how they derived these samples. The exact length
of the simulation was also decided by the modeling teams (Table
\@ref(tab:tableModels)). Further details on how each modeling team
prepared the simulations and the outputs are available in SM 2.

The simulations were run in the absence of management to a
simulated equilibrium ("Potential Natural Vegetation") with the current
climate, as the ultimate goal was to derive the regeneration niche. This
entails the assumption that (1) the observations from the forest
reserves reflect no traces of forest management and (2) there is an
equilibrium between forest dynamics and climate. While the former is
likely to be met at least in many of the EuFoRIA reserves, the latter
may be more debatable. However, in the absence of detailed data on the
history of each plot in the EuFoRIa network, some broad assumptions had
to be made. Both the width of the regeneration niche (i.e., in
environmental space) as well as the intensity of the ingrowth process
(i.e., the number of ingrowth trees per area and per unit of time) were
of interest.

The simulations were run for mixed species forests (not multiple
single-species simulations) using mixtures of eleven species or genera
for which ingrowth data of sufficient quality were available from
EuFoRIa: *Fagus sylvatica, Picea abies, Abies alba, Carpinus betulus,
Tilia cordata, Acer pseudoplatanus, Betula spp., Fraxinus excelsior,
Quercus spp., Alnus glutinosa, Pinus sylvestris.* Two models included
their standard set of species for the simulations which is much larger
(i.e., ForClim 1, ForClim 11). This also applies to TreeMig, but in this
model *Betula pendula* was not included because HERE **TREEMIG**
PROVIDES THE END OF THIS SENTENCE SAYING WHY B pendula WAS NOT INCLUDED.
In three models, fewer than these elveven species were simulated (4C,
xComp, LPJ-GUESS) (Table \@ref(tab:tableModels)). HERE **4C** PROVIDES
ONE SENTENCE WHY THEY SIMULATED LESS SPECIES. HERE **XCOMP** PROVIDES
ONE SENTENCE WHY THEY SIMULATED LESS SPECIES. HERE **LPJ-GUESS**
PROVIDES ONE SENTENCE WHY THEY SIMULATED LESS SPECIES. Finally, the
model aDGVM2 is based on a suite of broad Plant Functional Types rather
than individual species, and hence this model used its default approach
of PFTs for the simulations.

Each of the models reported the ingrowth number by sampling 200 times
over a 10-year interval for each species and per ha for each of the 200
sites. Multiple samples per site were used to better understand the
simulated variation within each site. This was done using different
strategies, depending on the model, including (1) sampling simulated
data from the same 1-ha plot in the equilibrium over time, or (2)
sampling several 1-ha plots from the simulated forest at one specific
point in time (in the equilibrium). This resulted in 880.000
observations per model that simulated the 11 species included in the
protocol (200 sites, 200 samples per site, 11 species and two diameter
thresholds). For the models that simulated additional species, their
ingrowth rates were aggregated as "others".

Three models did not provide results from all the simulated samples or
sites to avoid unrealistic results, as follows. In the model 4C, a
threshold of a maximum basal area of 100 $m^2 ha^{-1}$ was used to avoid
unrealistic stand basal area data, and therefore not all the samples and
sites were reported. The reason for this is that 4C is not suitable for
long-term simulations without management PLEASE **4C** VERIFY THIS IS
WHAT YOU DID. LPJ-GUESS had 2% of the sample outputs as grasslands that
were considered in the data as zero stand basal area and zero ingrowth
for all the tree species simulated. LPJ-GUESS could not simulate one of
the sites HERE **LPJ-GUESS** PROVIDES ONE SENTENCE WHY THEY DID NOT
PROVIDE ALL THE SITES. Landis-II missed 0.2% of the observations HERE
**LANDIS-II** PROVIDES ONE SENTENCE WHY THEY DID NOT PROVIDE ALL THE
SAMPLES/observations.

@@

The simulation results were analyzed regarding the ingrowth levels
(i.e., ingrowth number per unit time and space), species composition of
the ingrowth, and ingrowth mortality. The observed data were modeled
using a Generalized Additive Model with a negative binomial distribution
to better understand the trend and levels of the observed ingrowth,
relative to the simulation results per model.

@@

The ingrowth levels varied strongly across the 15 models and typically
did not match the levels found in the forest reserve data (Figure
\@ref(fig:over)). Ingrowth was overestimated in most models for both the
7 and 10 cm diameter thresholds, with the exception of the empirical
stand model SYBILA, the landscape model Landis-II and the global model
aDGVM2, which estimated ingrowth rates at the lower end of the
plausibility interval of the observed data (Figure \@ref(fig:over)). The
models with the largest overestimation were the stand model PICUS and
the landscape model TreeMig. For most models, the variability of
simulated ingrowth levels across the 200 sites (visible from the
interquartile range in the box plots of Figure \@ref(fig:over)) was
similar to or smaller than observed, with the notable exception of
PICUS, where simulated ingrowth variability was much larger than that of
any other model.

@@

In the observed data, ingrowth rates decreased clearly with increasing
stand basal area (Figure \@ref(fig:nvTrend710BA)). This trend was
captured by the Landis-II model only, albeit at much lower values of
basal area than in the observations. The other models featured
distinctly different trends, such as (1) an increase in ingrowth with
increasing total basal area followed by the absence of ingrowth at high
values of basal area (ForCEEPS, ForCEEPS(f), FORMIND, TreeMig and
LPJ-GUESS), (2) almost constant ingrowth levels with basal area (ForClim
11, xComp, LandClim and aDGVM2), or (3) an increase of ingrowth up to a
certain value of stand basal area followed by a decrease at even higher
values of stand basal area, with model-specific thresholds (4C, ForClim
1, PICUS, SYBILA and iLand).

@@

First, some models overestimated ingrowth across the entire gradients
for at least one of the five main tree species (Figure \@ref(fig:diff),
e.g., ForCEEPS, ForCEEPS(f), LandClim or LPJ-GUESS), while other models
overestimated ingrowth for more than one species for a large part of the
climate gradients, such as xComp, PICUS, SYBILA or LPJ-GUESS (cf. red
colors in Figure \@ref(fig:diff)). These trends were consistent for both
ingrowth thresholds, although the exact changes in the share of the
ingrowth basal area were somewhat different (Figure S7).

Second, there was some congruence in the simulation results by species
across the models. The share in the ingrowth by Silver fir (*Abies
alba*) was captured well across the gradient by most models except for
xComp, PICUS, LandClim and Landis-II, which overestimated the ingrowth
share of this species. There were also some peculiarities evident for
some models, such as 4C that did not simulate *Abies alba*. *Fagus
sylvatica* ingrowth was underestimated by many models across the
environmental gradients, including ForCEEPS(f) and FORMIND, which
featured negligible ingrowth at most sites for this species. Most of the
models overestimated the share of *Picea abies* ingrowth in the cold-wet
part of the gradients. *Pinus sylvestris* and *Quercus* spp. represented
a smaller share of the ingrowth in the observed data mostly at warm-dry
sites. Many models, however, erroneously featured ingrowth for these two
species along most of the climatic gradients although at a low basal
area share. Five models (4C, FORMIND, SYBILA, xComp, and LandClim) had
almost no ingrowth of *Pinus sylvestris* and *Quercus* spp., or did not
feature any ingrowth of these species at all.

@@

We recommend to further investigate the implications of the current
modeling strategies for tree establishment, particularly those related
to excess tree establishment and probably late excess mortality, by
focusing on a wider range of tree sizes and the related ingrowth
dynamics. We also recommend, especially for the landscape-level models,
the inclusion of explicit spatial considerations regarding tree ingrowth
[@beckage2003]; this, however, is facing a serious challenge regarding
appropriate data sets, as inventory data have wide coverage, but do not
allow for the assessment of spatial interaction effects.

>>>>

Most of the models overestimate ingrowth rates, which is only
compensated by some models with high simulated mortality rates in the
early stages of population dynamics. The effect of light availability on
ingrowth rates is captured better than the effect of temperature and
soil moisture, but patterns are not consistent across models. Simulated
species diversity of ingrowth matches the observed ranges quite well.
Models simulating higher species diversity at the stand level do not
feature higher ingrowth diversity. The ingrowth of *Abies alba* is
captured across temperature and soil moisture gradients, whereas
ingrowth of *Fagus sylvatica* is generally underestimated. *Picea abies*
ingrowth is generally overestimated, particularly at colder sites.
Increasing complexity in the tree establishment modules of the models is
not related to higher accuracy of simulated tree establishment.
Furthermore, individual model design is more important than scale
(stand, landscape, global) and approach (empirical, process-based) to
accurately capture tree establishment.

@@

In the present study, for the first time, a large number of forest
models commonly used to assess forest dynamics under climate change are
evaluated against a continental-scale, multi-species harmonized dataset
on tree establishment [@käber2022]. This is a challenging experiment for
these models as they do also have to estimate the forest stand above the
tree establishment. By tree establishment, we refer to the passing of
trees across a specific diameter threshold ("ingrowth"). We included
models that are based on a range of "philosophies", from models
operating at the stand to the global scale as well as the range of
models from empirically derived to "process-based" [e.g., @bugmann1996;
@fabrika2012; @mette2014; @lexer2001]. Because of the large variability
in tree establishment patterns in nature and the large number of factors
driving this process --- including some that are not incorporated in
most models such as deer browsing --- we do not aim at a detailed
statistical evaluation of each model. Instead, we aim at evaluating the
general ingrowth patterns and magnitudes simulated by the models, and
benchmark the simulated regeneration niche of multiple species against
empirical data along a wide environmental gradient of temperature,
moisture, and light availability.

@@

Fifteen models of forest dynamics (including two models featuring
variants) were used to assess simulated tree establishment. They
simulate forest dynamics at the stand, landscape or global scale (Table
\@ref(tab:tableModels)). The approaches used for model construction and
their origin differ strongly, with most of the models featuring a
largely "process-based" approach, whereas two models are based on
formulations derived from the statistical analysis of inventory data
(SIBYLA and xComp). Some of the models largely rely on the original
approach underlying forest gap models (e.g., ForClim 1 and LandClim),
whereas some are based on plant ecophysiological processes (e.g.,
FORMIND and iLand). The different approaches underlying the models have
strongly influenced the formulation of tree establishment.

The establishment modules can be categorized as 'regeneration' or
'recruitment' models [@vanclay1994; @konig_tree_2022] (see Table
\@ref(tab:tableModels)). Regeneration models include processes such as
flowering and pollination, seed production, seed dispersal, germination,
and seedling growth, which ultimately lead to the simulated number of
established trees. Recruitment models, in contrast, introduce a certain
number of new trees with certain characteristics such as biomass or
diameter, without explicitly considering earlier development processes.
We can further distinguish models that feature a feedback in the
establishment, i.e., where the level and species composition of
establishment is influenced by the existing adult tree community via the
production of seeds, seedlings or saplings, from models that do not
contain such a feedback.

The starting point for the models establishment differs as well, ranging
from seedbank, seed or seedling to sapling (i.e., trees much larger than
10 cm height, often at least 2-3 m tall). Models that start from seed
need to include a larger number of ecological processes such as
germination and survival of young seedlings, whereas models that started
from saplings have to aggregate via parameterizations several ecological
processes because they are not treated explicitly. This reduces model
complexity but comes at the cost of blurring process representation. In
the models used in this study the overall complexity in the
establishment modules varies considerably. Following @bugmann2022, we
can classify seven models (iLand, PICUS, LandClim, ForCEEPS, LPJ-GUESS,
ForClim and TreeMig) as having rather high complexity in their
establishment modules (mean establishment formulation complexity
complexity across all processes \>0, Table
\@ref(tab:)\@ref(tab:tableModels)).

Two models, ForClim and ForCEEPs, were used here with two alternative
variants. ForClim variant 1 [@bugmann1996] is based on a recruitment
module that adheres closely to the concept introduced by @botkin1972,
whereas ForClim variant 11 is adopting a rather different approach that
considers space filling along with different 'points of entry' for the
action of specific environmental factors [@huber2020]. These two model
variants allow us to evaluate a more process-based module (variant 11)
against a simple module (variant 1), while the rest of the model
structure is identical. Similarly, the two ForCEEPS variants allow us to
isolate the importance of the canopy feedback (i.e. simulated actual
composition and relative abundance of species in the plot)via seed trees
for the quantity and quality of simulated ingrowth, as one variant
includes this feedback whereas the other does not.

@@

Ingrowth data covering a wide range of environmental conditions are hard
to obtain, and this is one of the reasons why most models of forest
dynamics have never been confronted with a dataset covering such
gradients over a large number of sites, to evaluate how well ingrowth is
captured. The observations used here are derived from a novel and
unprecedented network of sites in forest reserves that represent the
range of environmental gradients in temperature and precipitation in
central Europe as compiled in the framework of the EuFoRIa network
[@euforia2019] (Figure \@ref(fig:map)). These forests are unmanged for
at least 10 years prior to their designation as reserves, and most of
them provide time series of natural forest dynamics over multiple
decades and up to ca. 80 years. The census periods range from 3 to 37
years, with an average of 14 years. The data provide information at the
tree level, thus allowing for sequential comparison such as
individual-tree ingrowth and death between the measurements.

We selected 200 sites from this network as the benchmarking dataset for
the simulation. The selection of sites is representative for the
environmental variation of the data. This was achieved by applying
k-means clustering to define 200 clusters of plots from the original set
of 869 plots along the environmental dimensions of temperature, water
balance, soil quality and exposition (i.e., slope and aspect). Each of
these sites featured at least two consecutive measurements. Ingrowth was
measured at threshold of 7 and 10 cm of stem diameter at breast height
in all sites except for 35 sites where only ingrowth at the 10 cm
diameter threshold was available. Plot size ranged from 0.02 ha to 5.52
ha, and the data were further processed and aggregated following
@käber2022 to provide ingrowth rates per ha and per decade. The
observations used in this study featured 30,900 established trees, and
ingrowth rates per site, sample and decade ranged from zero to 1246
trees per ha, with a mean of 56 trees per ha. For more details on this
unique dataset and the detailed establishment information, cf.
@käber2022.

@@

The overarching goal of the experiments was to assess the regeneration
niche as it arises from empirical data against its representation in a
wide range of models of forest dynamics. We define the regeneration
niche as the passing of a breast height diameter threshold of 7 or 10
cm, respectively (synonym: ingrowth). To this end, each modeling group
was provided with a detailed protocol (cf. Supplementary Material 1,
henceforth "SM") with instructions how to perform the simulations,
providing input variables on climate and soil conditions, and the
expected output variables. Neither were further site information (except
for the data specified below) nor any data on tree ingrowth provided
prior to the simulation. That is, the models were run in "blind flight"
mode.

The input variables were collected and aggregated to be adapted to the
needs of the different models. Time series of climatic variables from
1981 to 2018 were provided in hourly [Era5-land data, Muñoz Sabater
-@muñozsabater2019], daily and monthly resolution [CHELSA data, Karger
et al. -@karger2021]. Some variables required for some models such as
relative humidity or vapor pressure deficit had to be calculated from
these variables. The final strategy how to use the climate data was
different in each model, based on the approach that best suited the
model (Table \@ref(tab:tableModels)). Soil quality data was provided as
a continuous value between 1 and 5 [Soilgrids dataset, Hengl et al.
-@hengl2017]. The protocol also provided the elevation, slope and aspect
for each of the 200 sites [@asterscienceteam2019], but no other spatial
information such as coordinates was provided, with the exception of
iLand and aDGVM2, which used blurred coordinates to derive highly
detailed soil data.

The simulations were run in the absence of natural disturbances. Only
the model LPJ-GUESS had to include a background disturbance to increase
chances of shade-intolerant species to establish. The simulations were
set up to sample species-specific ingrowth rates per decade and per ha
in the equilibrium state of the model, typically entailing a "spin-up"
run (as we did not provide any forest data). The modeling teams decided
on the simulated area and how they derived these samples. The exact
length of the simulation was also decided by the modeling teams (Table
\@ref(tab:tableModels)). Further details on how each modeling team
prepared the simulations and the outputs are available in SM 2.

The simulations were run in the absence of management to a
simulated equilibrium ("Potential Natural Vegetation") with the current
climate, as the ultimate goal was to evaluate tree establishment under
comparable and near equilibrium conditions. This entails the assumption
that the observations from the forest reserves (1) reflect no traces of
forest management and (2) there is an equilibrium between forest
dynamics and climate. While the former might be starting to be visible
in many of the EuFoRIA reserves, the latter may be more debatable.
However, in the absence of detailed data on the history of each plot in
the EuFoRIa network, some broad assumptions had to be made. Both the
width of the regeneration niche (i.e., in environmental space) as well
as the intensity of the ingrowth process (i.e., the number of ingrowth
trees per area and per unit of time) were of interest.

The simulations were run for mixed species forests (not multiple
single-species simulations) using mixtures of eleven species or genera
for which ingrowth data of sufficient quality were available from
EuFoRIa: *Fagus sylvatica, Picea abies, Abies alba, Carpinus betulus,
Tilia cordata, Acer pseudoplatanus, Betula spp., Fraxinus excelsior,
Quercus spp., Alnus glutinosa, Pinus sylvestris.* The same set of
species were used in all the 200 sites. Two models included their
standard set of species for the simulations which is much larger (i.e.,
ForClim 1, ForClim 11, TreeMig). In three models, fewer than these
eleven species were simulated (4C, xComp, LPJ-GUESS) (Table
\@ref(tab:tableModels)). In 4C onl*y Fagus sylvatica*, *Picea abies*,
*Betula* spp., *Quercus* spp. and *Pinus sylvestris* are parameterised.
XComp simulations did not consider *Fraxinus excelsior* due to a
depreciated species parameterisation. In LPJ-GUESS, *Acer
pseudoplatanus* and *Alnus glutinosa* are not parameterized and
therefore these species could not be included in the simulation.
Finally, the model aDGVM2 does not simulate individual species. Rather,
community assembly processes and trait filtering generate plant
communities that are adjusted to the biotic and abiotic conditions and
plants can be classified into different ecological strategies based on
their trait values in a post-processing step.

Each of the models reported the ingrowth number by sampling 200 times
over a 10-year interval for each species and per ha for each of the 200
sites. Multiple samples per site were used to better understand the
simulated variation within each site. This was done using different
strategies, depending on the model, including (1) sampling simulated
data from the same 1 ha plot in the equilibrium over time, (2) sampling
several 1 ha plots from the simulated forest at one specific point in
time (in the equilibrium) or (3) a combination of (1) and (2) . This
resulted in 880,000 observations per model that simulated the 11 species
included in the protocol (200 sites, 200 samples per site, 11 species
and two diameter thresholds). For the models that simulated additional
species, their ingrowth rates were aggregated as "others".

Three models did not provide results from all the simulated samples or
sites to avoid unrealistic results, as follows. In the model 4C, a
threshold of a maximum basal area of 90 $m^2 ha^{-1}$ was used to avoid
unrealistic stand basal area data, and therefore not all the samples and
sites were reported. The reason for this is that 4C is not suitable for
long-term simulations without management, due to misrepresentations in
density-dependent mortality processes in long-term simulations and
assumptions of tree geometry that lead to unrealistic single tree
dimensions of very old individuals. LPJ-GUESS had 2% of the sample
outputs as grasslands that were considered in the data as zero stand
basal area and zero ingrowth for all the tree species simulated and one
site was not simulated.

@@

The simulation results were analyzed regarding the ingrowth levels
(i.e., ingrowth number per unit time and space), species composition of
the ingrowth, and ingrowth mortality. The observed data were modeled
using a Generalized Additive Model (GAM) with a negative binomial
distribution to better understand the trend and levels of the observed
ingrowth, relative to the simulation results per model. To avoid
overfitting we used the restricted maximum likelihood

@@

The ingrowth levels varied strongly across the 15 models and typically
did not match the levels found in the forest reserve data (Figure
\@ref(fig:over)). Ingrowth was overestimated in most models for both the
7 and 10 cm diameter thresholds, with the exception of the empirical
stand model SIBYLA, the landscape model Landis-II and the global model
aDGVM2, which estimated ingrowth rates at the lower end of the
plausibility interval of the observed data (Figure \@ref(fig:over)). The
models with the largest overestimation were the stand model PICUS and
the landscape model TreeMig. For most models, the variability of
simulated ingrowth levels across the 200 sites (visible from the
interquartile range in the box plots of Figure \@ref(fig:over)) was
similar to or smaller than observed, with the notable exception of
PICUS, where simulated ingrowth variability was much larger than that of
any other model.

@@

In the observed data, ingrowth rates decreased clearly with increasing
stand basal area (Figure \@ref(fig:nvTrend710BA)). This trend was
captured by the Landis-II model only, albeit at much lower values of
basal area than in the observations. The other models featured
distinctly different trends, such as (1) an increase in ingrowth with
increasing total basal area followed by the absence of ingrowth at high
values of basal area (ForCEEPS, ForCEEPS(f), FORMIND, TreeMig and
LPJ-GUESS), (2) almost constant ingrowth levels with basal area (ForClim
11, xComp, LandClim and aDGVM2), or (3) an increase of ingrowth up to a
certain value of stand basal area followed by a decrease at even higher
values of stand basal area, with model-specific thresholds (4C, ForClim
1, PICUS, SIBYLA and iLand).

@@

First, some models overestimated ingrowth across the entire gradients
for at least one of the five main tree species (Figure \@ref(fig:diff),
e.g., ForCEEPS, ForCEEPS(f), LandClim or LPJ-GUESS), while other models
overestimated ingrowth for more than one species for a large part of the
climate gradients, such as xComp, PICUS, SIBYLA or LPJ-GUESS (cf. red
colors in Figure \@ref(fig:diff)). These trends were consistent for both
ingrowth thresholds, although the exact changes in the share of the
ingrowth basal area were somewhat different (Figure S7).

Second, there was some congruence in the simulation results by species
across the models. The share in the ingrowth by Silver fir (*Abies
alba*) was captured well across the gradient by most models except for
xComp, PICUS, LandClim and Landis-II, which overestimated the ingrowth
share of this species. There were also some peculiarities evident for
some models, such as 4C that did not simulate *Abies alba*. *Fagus
sylvatica* ingrowth was underestimated by many models across the
environmental gradients, including ForCEEPS(f) and FORMIND, which
featured negligible ingrowth at most sites for this species. Most of the
models overestimated the share of *Picea abies* ingrowth in the cold-wet
part of the gradients. *Pinus sylvestris* and *Quercus* spp. represented
a smaller share of the ingrowth in the observed data mostly at warm-dry
sites. Many models, however, erroneously featured ingrowth for these two
species along most of the climatic gradients although at a low basal
area share. Five models (4C, FORMIND, SIBYLA, xComp, and LandClim) had
almost no ingrowth of *Pinus sylvestris* and *Quercus* spp., or did not
feature any ingrowth of these species at all.

@@

We recommend to further investigate the implications of the current
modeling strategies for tree establishment, particularly those related
to excess tree establishment and probably late excess mortality, by
focusing on a wider range of tree sizes and the related ingrowth
dynamics. We also recommend, especially for the landscape-level models,
the inclusion of explicit spatial considerations regarding tree ingrowth
[@beckage2003]; this, however, is facing a serious challenge regarding
appropriate datasets, as inventory data have wide coverage, but do not
allow for the assessment of spatial interaction effects.
